---
title: "Group 1 — Yogurt Study Data Exploration"
author: "Workshop Participant"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Overview

This document explores the **Yogurt study dataset** (Group A / Group 1).
The study investigates how eating yogurt affects the vaginal microbiome.
Participants were split into two **arms**: a **yogurt** group and an
**unchanged diet** group. Samples were taken at two **time points**:
baseline and after antibiotics.

The data is split across four CSV files, linked by participant IDs
(`pid`) and sample IDs (`sample_id`).

---

## Load libraries

```{r}
#| message: false
library(tidyverse)
library(knitr)
```

---

## File 1 — Sample IDs (`00_sample_ids_yogurt.csv`)

Each row is **one sample** from one participant at one time point.

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID (e.g. `pid_25`). Links to the metadata file. |
| `time_point` | When the sample was taken — `baseline` (before study) or `after_antibiotic` (after a course of antibiotics). |
| `arm` | Which group the participant was assigned to — `yogurt` or `unchanged_diet`. |
| `sample_id` | Unique lab sample identifier (e.g. `YOG002`). Links to qPCR and Luminex results. |

```{r}
sample_ids <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")

glimpse(sample_ids)
```

```{r}
# Count samples by time point and arm
sample_ids |> 
  count(time_point, arm)
```

---

## File 2 — Participant Metadata (`01_participant_metadata_yogurt.csv`)

Each row is **one participant** with their demographic and health info.

| Column | Meaning |
|--------|---------|
| `pid` | Participant ID. Links to the sample IDs file. |
| `arm` | Which group — `yogurt` or `unchanged_diet`. |
| `days_since_last_sex` | Days since the participant last had sex (integer). |
| `birth_control` | Type of birth control — `no hormonal birth control` or `Depoprovera` (a hormonal injectable). |
| `age` | Age in years. |
| `education` | Education level (self-reported, categorical). |
| `sex` | Sex of the participant (all recorded as `female`). |

```{r}
metadata <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")

glimpse(metadata)
```

```{r}
# Age distribution by arm
ggplot(metadata, aes(x = arm, y = age, fill = arm)) +
  geom_boxplot() +
  labs(title = "Age distribution by study arm", y = "Age (years)", x = NULL) +
  theme_minimal()
```

```{r}
# Birth control by arm
metadata |> 
  count(arm, birth_control) |> 
  kable()
```

---

## File 3 — qPCR Results (`02_qpcr_results_yogurt.csv`)

Each row is **one sample** with qPCR measurements of bacterial DNA.
qPCR (quantitative PCR) is a lab technique that measures how much of a
specific bacterium's DNA is present in a sample.

| Column | Meaning |
|--------|---------|
| `sample_id` | Lab sample ID. Links to the sample IDs file. |
| `qpcr_bacteria` | Amount of **total bacterial DNA** detected. |
| `qpcr_crispatus` | Amount of *Lactobacillus crispatus* DNA (a "good" bacteria linked to vaginal health). |
| `qpcr_iners` | Amount of *Lactobacillus iners* DNA (another *Lactobacillus* species, considered less protective). |

```{r}
qpcr <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")

glimpse(qpcr)
```

```{r}
# Join qPCR results with sample info so we can compare arms
qpcr_with_arm <- sample_ids |> 
  left_join(qpcr, by = "sample_id")

# Compare total bacteria by arm and time point
ggplot(qpcr_with_arm, aes(x = arm, y = qpcr_bacteria, fill = arm)) +
  geom_boxplot() +
  facet_wrap(vars(time_point)) +
  scale_y_log10() +
  labs(title = "Total bacterial load by arm and time point",
       y = "Total bacteria (log scale)", x = NULL) +
  theme_minimal()
```

```{r}
# Compare L. crispatus vs L. iners by arm
# Higher crispatus / lower iners is generally considered healthier
qpcr_with_arm |> 
  pivot_longer(cols = c(qpcr_crispatus, qpcr_iners),
               names_to = "bacteria",
               values_to = "count") |> 
  ggplot(aes(x = arm, y = count + 1, fill = bacteria)) +
  geom_boxplot() +
  facet_wrap(vars(time_point)) +
  scale_y_log10() +
  labs(title = "L. crispatus vs L. iners by arm and time point",
       y = "Bacterial DNA amount (+1, log scale)", x = NULL) +
  theme_minimal()
```

---

## File 4 — Luminex Results (`03_luminex_results_yogurt.csv`)

Each row is **one cytokine measurement** for one sample. This is "long"
format — each sample appears in multiple rows, one per cytokine measured.
Luminex is a lab technique that measures multiple proteins (cytokines) at
once. Cytokines are signalling molecules that the immune system uses to
communicate — higher levels can mean more inflammation.

| Column | Meaning |
|--------|---------|
| `sample_id` | Lab sample ID. Links to the sample IDs file. |
| `cytokine` | Which cytokine was measured (e.g. `IL-1a`, `TNFa`, `IL-8`). 10 different cytokines in total. |
| `conc` | Concentration of that cytokine in the sample (units as provided by the lab). |
| `limits` | Whether the measurement was `within limits` (reliable) or `out of range` (below the test's detection limit). |

```{r}
luminex <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")

glimpse(luminex)
```

```{r}
# List all cytokines measured
unique(luminex$cytokine)
```

```{r}
# How many measurements are within vs out of range?
luminex |> 
  count(limits)
```

```{r}
# Join Luminex results with sample info
luminex_with_arm <- sample_ids |> 
  left_join(luminex, by = "sample_id")

# Look at a few key inflammatory cytokines across arms
key_cytokines <- c("IL-1b", "IL-6", "IL-8", "TNFa")

luminex_with_arm |> 
  filter(cytokine %in% key_cytokines, limits == "within limits") |> 
  ggplot(aes(x = arm, y = conc, fill = arm)) +
  geom_boxplot() +
  facet_grid(cytokine ~ time_point, scales = "free_y") +
  labs(title = "Key inflammatory cytokines by arm and time point",
       y = "Concentration", x = NULL) +
  theme_minimal()
```
